### 2012-10-16 doc/prelim/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished rng traits and boost implementations and tests
working on adjoint section per spaniers work
incrporating more of Pauls changes. still a few more to go

\chapter{Stochastic Solution Methods for Nonlinear Problems}
\label{ch:nonlinear_problem}
Nonlinear equation sets are a common occurrence in multiphysics
problems. Systems of partial differential equations such as those that
describe fluid flow or more general transport processes when
discretized by conventional methods yield discrete sets of stiff
equations with nonlinearities present in the variables. Traditionally,
such systems have been solved by linearizing them in a form where the
nonlinearities in the variables are eliminated and more traditional
linear methods can be used for solutions. Often characterized as
segregated methods where physics operators are split and their action
on the system approximated in steps, such methods lack consistency and
accuracy in resolving the nonlinear component of the solution. In the
last 30 years, fully consistent nonlinear methods based on Newton's
method have become more popular and many advances have been made in
the computational physics field to employ these methods.

Currently, Krylov methods are used with Newton's method to form a
robust solution sequence for these types of problems. However, given
our goals of advancing stochastic solver technology for linear
problems, we aim to apply them to these fully consistent schemes in
order to gauge if some improvements can be made. We therefore devise a
new method using MCSA in conjunction with new methods, accommodating the
fact that MCSA requires an explicitly formed linear operator in its
implementation as compared to Krylov methods. In addition to handling
this dependency, we will construct a parallel strategy for a nonlinear
MCSA-based method.

>>>>

\chapter{Monte Carlo Solution Methods for Nonlinear Systems}
\label{ch:nonlinear_problem}
Nonlinear equation sets are a common occurrence in multiphysics
problems. Systems of partial differential equations such as those that
describe fluid flow or more general transport processes when
discretized by conventional methods yield discrete sets of stiff
equations with nonlinearities present in the variables. Traditionally,
such systems have been solved by linearizing them in a form where the
nonlinearities in the variables are eliminated and more traditional
linear methods can be used for solutions. Often characterized as
segregated methods where physics operators are split and their action
on the system approximated in steps, such methods lack consistency and
accuracy in resolving the nonlinear component of the solution. In the
last 30 years, fully consistent nonlinear methods based on Newton's
method have become more popular and many advances have been made in
the computational physics field to employ these methods.

In the context of solving standalone linear systems, Monte Carlo
methods do not provide significant merit over Krylov methods due to
the fact that the linear operator must be explicitly formed. For many
applications, such a requirement is prohibitive and perhaps not even
feasible to implement. Therefore, a Monte Carlo solver is best suited
for situations in which not only are Krylov methods applicable, but
also in which the operator is readily, if not naturally,
formed. Modern nonlinear methods meet both of these requirements with
Newton methods used in conjuction with Krylov methods for a robust,
fully implicit solution strategy. Furthermore, modern techniques exist
that permit the automatic construction of the linear operator
generated within a Newton method based on the nonlinear residual
evaluations, providing all of the components necessary for a Monte
Carlo solver to provide value. We therefore devise a new nonlinear
method based on the MCSA algorithm and Newton's method and discuss its
potential benefits.

### 2012-10-16 doc/prelim/document/chapters/parallel_mc/parallel_mc.tex
incrporating more of Pauls changes. still a few more to go

\section{Preliminaries}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can using diagonal splitting in a similar manner
as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
define the following operator\footnote{It should be noted that
  non-diagonal splittings have been recently explored in
  \citep{srinivasan_monte_2010} and have the potential to improve
  efficiency.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this inverse sum to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation will therefore lead to an
approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering its $i^{th}$
component:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is a particular random walk sequence
permutation. We can generate these sequences of transitions through
Monte Carlo random walks by assigning them both a probability and
weight. As a reinterpretation of the iteration matrix, we then form
the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}).

@@

As we seek only approximate solutions, we need only to perform a
minimal number of random walks in order to generate an approximation
for $\ve{x}$. If we are only to approximate the solution, we also need
conditions by which we may terminate a random walk. We do this by
noticing that the factors added to
Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
small due to their definition in Eq~(\ref{eq:direct_weight}) and
therefore their contributions to the solution estimate will become
negligible. Using this, we choose terminate a random walk sequence with
a \textit{weight cutoff}, $W_c$, that is enforced when $W_m < W_c$ for
a particular random walk permutation.

@@

\subsection{Preconditioning MCSA}
\label{sec:stochastic_preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. To achieve this, a Jacobi
preconditioner, a form of left preconditioning, is used such that the
preconditioning matrix $\ve{M}$ is:
\begin{equation}
  \ve{M} = diag(\ve{A})\:,
  \label{eq:jacobi_preconditioner}
\end{equation}
such that its application means we are instead solving the following
linear system:
\begin{equation}
  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
  \label{eq:jacobi_precond_linear_problem}
\end{equation}
Next, we can apply MCSA to solve
Eq~(\ref{eq:jacobi_precond_linear_problem}): 
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
    \ve{x}^{k+1/2}\:.
    \label{eq:jacobi_preconditioned_mcsa}
  \end{gather}
\end{subequations}
Choosing Jacobi preconditioning with MCSA is advantageous for several
reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for
all $\ve{A}$ and is easy to formulate as the inversion of $\ve{M}$ is
trivial. Second, because the adjoint Monte Carlo method used within MCSA to
compute the correction operates on a linear problem with the
preconditioned operator, then $\ve{H}$ in the adjoint solver will have
a zero term in each of its diagonal elements, thereby eliminating
all in-state transitions during the random walk sequence. Because of
this, Jacobi preconditioning should always be performed, regardless of
any other preconditioning that is applied to the system.

@@

In the context of radiation transport, Brunner and colleagues provided
a survey of algorithms for achieving this as implemented in production
implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
work they identify two data sets that are required to be communicated:
the sharing of particles that are transported from one domain to
another and therefore from one processor to another and a global
communication that signals if particle transport has been completed on
all processors. The algorithms presented are a fully-locking
synchronous scheme, an asynchronous-send/synchronous-receive pattern,
a traditional master/slave scheme, and a modified master/slave scheme
the implements a binary tree pattern for the global reduction type
operations needed to communicate between the master and slave
processes. They observed that the modified master/slave scheme
performed best in that global communications were implemented more
efficiently than those required by the asynchronous
scheme. Furthermore, none of these schemes handled load-imbalanced
cases efficiently. Such cases will be common if the source sampled in
the Monte Carlo random walk is not isotropic and evenly distributed
throughout the global domain. It was noted that efficiencies were
improved by increasing the frequency by which particle data was
communicated between domain-adjacent processors. However, this
ultimately increases communication costs. In 2009, Brunner extended
his work by using a more load-balanced approach with a fully
asynchronous communication pattern
\citep{brunner_efficient_2009}. Although the extended implementation
was more robust and allowed for scaling to larger numbers of
processors, performance issues were still noted with parallel
efficiency improvements needed in both the weak and strong scaling
cases for unbalanced problems. These results led Brunner to conclude
that a combination of domain decomposition and domain replication
could be used to solve some of these issues.

\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsubsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed the similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its subdomains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication only occurs within a set, limiting communications
during the transport procedure to a group of $O(100)$ processors, a
number that was shown to have excellent parallel efficiencies in
Brunner's work and therefore will scale well in this algorithm. The
overlapping domains within each set also demonstrated reduced
communication costs. On each processor, the source is sampled in the
local domain that would exist if no overlap was used while tallies can
be made over the entire overlapping domain.

@@

Procassini's solutions to these problems are elegant in that they
require a minimal amount of modification to be applied to the Monte
Carlo algorithm. To solve the first issue, in order to ensure each
particle maintains an invariant random number sequence that determines
its behavior regardless of domain decomposition, each particle is a
random number seed that describes its current state upon entering the
domain of a new processor. These seeds are derived from the actual
geometric location of the particle such that it is decomposition
invariant. Non-commutative floating point operations are solved by
instead mapping floating point values to 64-bit integer values for
which additions will always be commutative. Once the operations are
complete, these integers are mapped back to floating point values.

>>>>

\section{Preliminaries}
\label{sec:mc_preliminaries}
We begin our discussion of Monte Carlo methods using these texts by
seeking a solution to Eq~(\ref{eq:linear_problem}). For a given linear
operator $\ve{A}$, we can use diagonal splitting in a similar manner
as the stationary method in Eq~(\ref{eq:linear_split_equation2}) to
define the following operator\footnote{It should be noted that
  non-diagonal splittings have been recently explored in
  \citep{srinivasan_monte_2010} and have the potential to improve
  efficiency.}:
\begin{equation}
  \ve{H} = \ve{I} - \ve{A}\:,
  \label{eq:linear_mc_iteration_matrix}
\end{equation}
such that we are solving the system:
\begin{equation}
  \ve{x} = \ve{H} \ve{x} + \ve{b}\:.
  \label{eq:richardson_split}
\end{equation}
We can then form an alternative representation for $\ve{A}^{-1}$ by
generating the \textit{Neumann series}:
\begin{equation}
  \ve{A}^{-1} = (\ve{I}-\ve{H})^{-1} = \sum_{k=0}^{\infty} \ve{H}^k\:,
  \label{eq:neumann_series}
\end{equation}
which will converge if the spectral radius of $\ve{H}$ is less than
1. If we then apply this inverse sum to the right hand side of
Eq~(\ref{eq:linear_problem}) we acquire the solution to the linear
problem:
\begin{equation}
  \ve{A}^{-1}\ve{b} = \sum_{k=0}^{\infty} \ve{H}^k\ve{b} = \ve{x}\:.
  \label{eq:neumann_solution}
\end{equation}
An approximation of this summation will therefore lead to an
approximation of the solution. If we expand the summation with a
succession of matrix-vector multiply operations, we arrive at an
alternative perspective of this summation by considering its $i^{th}$
component:
\begin{equation}
  x_i = \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
  \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:,
  \label{eq:expanded_neumann_solution}
\end{equation}
which can interpreted as a series of transitions between states,
\begin{equation}
 \nu = i \rightarrow i_1 \rightarrow \cdots \rightarrow i_{k-1}
 \rightarrow i_{k}\:,
  \label{eq:mc_walk_permutation}
\end{equation}
in $\ve{H}$ where $\nu$ is a particular random walk sequence
permutation. We can generate these sequences of transitions through
Monte Carlo random walks by assigning them both a probability and
weight. As a reinterpretation of the iteration matrix, we then form
the \textit{Neumann-Ulam decomposition} of \ve{H}:
\begin{equation}
  \ve{H} = \ve{P} \circ \ve{W}\:,
  \label{eq:neumann_ulam_decomposition}
\end{equation}
where $\circ$ denotes the Hadamard product operation\footnote{The
  Hadamard product $\ve{A} = \ve{B} \circ \ve{C}$ is defined
  element-wise as $a_{ij} = b_{ij} c_{ij}$.}, $\ve{P}$ denotes the
transition probability matrix, and $\ve{W}$ denotes the transition
weight matrix. This decomposition, a generalization of Dimov's work
\citep{dimov_new_1998}, is an extension of the original Neumann-Ulam
scheme in that now a weight cutoff can be used to terminate a random
walk sequence. The formulation of $\ve{P}$ and $\ve{W}$ will be
dependent on whether we choose a direct or adjoint Monte Carlo
sequence to estimate the state transitions in
Eq~(\ref{eq:expanded_neumann_solution}).

@@

In cases where we seek only approximate solutions, we need only to
perform a minimal number of random walks in order to generate an
approximation for $\ve{x}$. If we are only to approximate the
solution, we also need conditions by which we may terminate a random
walk. We do this by noticing that the factors added to
Eq~(\ref{eq:direct_permutation_weight}) will become diminishingly
small due to their definition in Eq~(\ref{eq:direct_weight}) and
therefore their contributions to the solution estimate will become
negligible. Using this, we choose terminate a random walk sequence
with a \textit{weight cutoff}, $W_c$, that is enforced when $W_m <
W_c$ for a particular random walk permutation.

@@

\subsection{Preconditioning MCSA}
\label{sec:stochastic_preconditioning}
In most cases, at least a minimal amount of \textit{preconditioning}
of the linear system will be required in order to use the class of
stochastic methods described. Although these methods have no symmetry
requirements for convergence, they do require that the spectral radius
of the iteration matrix be less than one. To achieve this, a Jacobi
preconditioner, a form of left preconditioning, is used such that the
preconditioning matrix $\ve{M}$ is:
\begin{equation}
  \ve{M} = diag(\ve{A})\:,
  \label{eq:jacobi_preconditioner}
\end{equation}
such that its application means we are instead solving the following
linear system:
\begin{equation}
  \ve{M}^{-1}\ve{A}\ve{x} = \ve{M}^{-1}\ve{b}\:.
  \label{eq:jacobi_precond_linear_problem}
\end{equation}
Next, we can apply MCSA to solve
Eq~(\ref{eq:jacobi_precond_linear_problem}): 
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = (\ve{I} - \ve{M}^{-1}\ve{A})\ve{x}^k +
    \ve{M}^{-1}\ve{b}\:,\\ \ve{r}^{k+1/2} = \ve{M}^{-1}\ve{b} -
    \ve{M}^{-1}\ve{A}\ve{x}^{k+1/2}\:,\\ \ve{M}^{-1}\ve{A}\delta\ve{x}^{k+1/2}
    = \ve{r}^{k+1/2}\:,\\ \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta
    \ve{x}^{k+1/2}\:.
    \label{eq:jacobi_preconditioned_mcsa}
  \end{gather}
\end{subequations}
Choosing Jacobi preconditioning with MCSA is advantageous for several
reasons. First, $\rho(\ve{I} - \ve{M}^{-1}\ve{A}) < 1$ is true for all
$\ve{A}$ and is easy to formulate because the inversion of $\ve{M}$ is
trivial. Second, because the adjoint Monte Carlo method used within
MCSA to compute the correction operates on a linear problem with the
preconditioned operator, then $\ve{H}$ in the adjoint solver will have
a zero term in each of its diagonal elements, thereby eliminating all
in-state transitions during the random walk sequence. Because of this,
Jacobi preconditioning should always be performed, regardless of any
other preconditioning that is applied to the system.

@@

In the context of radiation transport, Brunner and colleagues provided
a survey of algorithms for achieving this as implemented in production
implicit Monte Carlo codes \citep{brunner_comparison_2006}. In their
work they identify two data sets that are required to be communicated:
the sharing of particles that are transported from one domain to
another and therefore from one processor to another and a global
communication that signals if particle transport has been completed on
all processors. The algorithms presented are a fully-locking
synchronous scheme, an asynchronous-send/synchronous-receive pattern,
a traditional master/slave scheme, and a modified master/slave scheme
the implements a binary tree pattern for the global reduction type
operations needed to communicate between the master and slave
processes. They observed that the modified master/slave scheme
performed best in that global communications were implemented more
efficiently than those required by the asynchronous
scheme. Furthermore, none of these schemes handled load-imbalanced
cases efficiently. Such cases will be common if the source sampled in
the Monte Carlo random walk is not isotropic and not evenly
distributed throughout the global domain. It was noted that
efficiencies were improved by increasing the frequency by which
particle data was communicated between domain-adjacent
processors. However, this ultimately increases communication costs. In
2009, Brunner extended his work by using a more load-balanced approach
with a fully asynchronous communication pattern
\citep{brunner_efficient_2009}. Although the extended implementation
was more robust and allowed for scaling to larger numbers of
processors, performance issues were still noted with parallel
efficiency improvements needed in both the weak and strong scaling
cases for unbalanced problems. These results led Brunner to conclude
that a combination of domain decomposition and domain replication
could be used to solve some of these issues.

\subsubsection{Multiple-Set Overlapping-Domain Decomposition}
\label{subsubsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed the similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its subdomains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

Procassini's solutions to these problems are elegant in that they
require a minimal amount of modification to be applied to the Monte
Carlo algorithm. To solve the first issue, in order to ensure each
particle maintains an invariant random number sequence that determines
its behavior regardless of domain decomposition, each particle is
assigned a random number seed that describes its current state upon
entering the domain of a new processor. These seeds are derived from
the actual geometric location of the particle such that it is
decomposition invariant. Non-commutative floating point operations are
overcome by instead mapping floating point values to 64-bit integer
values for which additions will always be commutative. Once the
operations are complete, these integers are mapped back to floating
point values.
